---
title: "Exploratory Data Analysis and Modeling: Milestone Report"
author: "Prabhat Kumar"
date: "18 March 2016"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

# Executive Summary
The goal of this project is just to display that I've gotten used to working with the [SwiftKey, Inc.](https://swiftkey.com/en/) data and that I'm on track to create the prediction algorithm. I will explain my exploratory analysis and my goals for the eventual app and algorithm.

The objective of this project is to: (1). Demonstration of how to download the data and have successfully loaded it in, (2). Create a basic report of summary statistics about the data sets, (3). Report any interesting findings that I amassed so far, and (4). Get feedback on my plans for creating a prediction algorithm and Shiny app.

# Data Acquisition
I have acquired the data from given URL, [Coursera-SwiftKey.zip](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip). Further, unzipped the data in to ```data``` directory, by help of bash command ```unzip``` <i>i.e.</i>, is used for opening the archive; ```unzip("./data/Coursera-SwiftKey.zip")```.

```{r Environment Cleaning.}
# Remove Objects from a Specified Environment.
rm(list = ls())
```

# Reading the SwiftKey Data
I'm using here ```readLines``` method, with ```encoding = "UTF-8", skipNul = TRUE``` parameters to load/read en_US data sets, <i>i.e.</i>, en_US.blogs.txt/en_US.news.txt/en_US.twitter.txt.

```{r Reading the SwiftKey Data.}
enBlogs   <- readLines("./data/final/en_US/en_US.blogs.txt",   encoding = "UTF-8", skipNul = TRUE)
enNews    <- readLines("./data/final/en_US/en_US.news.txt",    encoding = "UTF-8", skipNul = TRUE)
enTwitter <- readLines("./data/final/en_US/en_US.twitter.txt", encoding = "UTF-8", skipNul = TRUE)
```
